---
title: "Neighborhood Built Environment Profiles in Bogotá"
subtitle: "Spatial visualization"
author: "Heli Xu"
date: 2/12/24
format: 
  html:
    code-fold: true
    toc: true
    toc-depth: 4
    self-contained: true
editor: visual
---

After using finite mixture modeling to [cluster the neighborhoods (ZATs)](../ZAT_profile/ZAT_profile.qmd) across all the ZAT-level indicators on road infrastructures, we'll visualize the resulting neighborhood profiles on maps.

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(sf)
library(tmap)
library(ggplot2)
```

## Data preparation

As mentioned in the previous post, all ZAT-level indicators were categorized into two domains, street design and transportation, and the neighborhood profiles are generated for each ZAT unit from the Gaussian mixture model-based clustering across the variables in each domain, as well as in both domains together. From the modeling results, we could extract the domain-specific and overall cluster information of all ZATs with `clusters()`, and join the data with the ZAT-level shapefiles in `ZAT folder` to get the geometry.

```{r, eval=FALSE}
zat_cluster <- zat_std2 %>% 
  select(ZAT) %>% 
  mutate(all = clusters(all_mix_c),
         street = clusters(st_mix_c),
         transp = clusters(tp_mix_c)) %>% 
  left_join(zat, by = "ZAT") %>% 
  select(ZAT, all, street, transp, geometry) %>% 
  st_as_sf()
```

```{r, echo=FALSE}
zat_cluster <- readRDS("../../../../clean_data/zat_cluster.rds")
```

## Mapping the profiles

To spatially visualize the neighborhood profiles, we'll create 3 maps to plot the cluster assignment of each ZAT unit for street design domain, transportation domain and both domains together.

```{r, warning=FALSE}
street <- zat_cluster %>%
  select(ZAT, geometry, street) %>%
  drop_na() %>%
  #using NAD83 in tmap
  tm_shape(projection = sf::st_crs(26915))+
  tm_polygons("street",
    palette = "YlGn",
    style = "cat",
    title = "cluster")+
  tm_layout(panel.show = TRUE,
    panel.labels = "Street Design Profiles",
    panel.label.fontface = "bold",
    title.position = c("left", "TOP"),
    legend.position = c("RIGHT", "bottom"),
    legend.title.size = 0.9,
    legend.width = 2)

transp <- zat_cluster %>%
  select(ZAT, geometry, transp) %>%
  drop_na() %>%
  #using NAD83 in tmap
  tm_shape(projection = sf::st_crs(26915))+
  tm_polygons("transp",
    palette = "YlGn", 
    style = "cat",
    title = "cluster")+
  tm_layout(panel.show = TRUE,
    panel.labels = "Transportation Profiles",
    panel.label.fontface = "bold",
    title.position = c("left", "TOP"),
    legend.position = c("RIGHT", "bottom"),
    legend.title.size = 0.9,
    legend.width = 2)

all_ind <- zat_cluster %>%
  select(ZAT, geometry, all) %>%
  drop_na() %>%
  #using NAD83 in tmap
  tm_shape(projection = sf::st_crs(26915))+
  tm_polygons("all",
    palette = "YlGn", 
    style = "cat",
    title = "cluster")+
  tm_layout(panel.show = TRUE,
    panel.labels = "All-Indicator Profiles",
    panel.label.fontface = "bold",
    title.position = c("left", "TOP"),
    legend.position = c("RIGHT", "bottom"),
    legend.title.size = 0.9,
    legend.width = 2)

plots <- list(street, transp, all_ind)
  
current.mode <- tmap_mode("plot")

tmap_arrange(
    plots,
    nrow = 1,
    width = c(0.34, 0.33, 0.33)
  )
```

We can see that the clustering by street design and/or transportation domain(s) yielded a relatively consistent pattern of neighborhood profiles in Bogotá.

## Assess variable influence

Since these profiles are modeled across the variables (in each domain), we'll now explore further which variables may be more influential in the clustering result, which will also be important in our interpretation of the different profiles/clusters.

One way to do that is to visualize each variable by the ZAT unit and see which variables have a better cluster separation (as in the previous post). In addition, we can examine the mean and variance of each variable see which ones are significantly different across clusters.

### Street Design Domain

```{r}
zat_cluster_var <- zat_cluster %>%
  as.data.frame() %>% 
  select(-geometry) %>% 
  left_join(zat_std2, by = "ZAT")

street_dm <- c("st_4ln_length_log", "bikelane_m_log", "trlight_per_km2", 
            "sttree_per_km2", "bridg_per_km2")

transportation_dm <- c("BUSTOPDENS", "bus_length_log", "brt_length_log", "numrbp_per_km2")

st_var <- zat_cluster_var %>% 
  select(-ZAT, -transp, -all) %>% 
  pivot_longer(-street, names_to = "indicator") %>% 
  filter(indicator %in% street_dm) %>% 
  group_by(street, indicator) %>% 
  mutate(mean = mean(value),
    sd = sd(value))

ggplot(st_var, aes(x= factor(street), y = value, fill = factor(street)))+
  geom_boxplot() +
  scale_fill_brewer(palette = "YlGn") + # Change fill colors
  theme_minimal() + # Use a minimal theme
  labs(title = "Street Design Indicators by Neighborhood Profiles", 
    x = "Cluster", 
    y = "Value",
    fill = "cluster") +
  facet_wrap(~indicator, scales = "free", nrow = 1)
```

It looks like the ... has the most separation between clusters and therefore is the most influential variable in this model-based clustering result. For this variable, the map looks like this,

and it looks like the cluster 1 neighborhoods is corresponding to the neighborhoods that have more/less street.

### Transportation Domain

```{r}
transp_var <- zat_cluster_var %>% 
  select(-ZAT, -street, -all) %>% 
  pivot_longer(-transp, names_to ="indicator") %>% 
  filter(indicator %in% transportation_dm) %>% 
  group_by(transp, indicator) %>% 
  mutate(mean = mean(value),
         sd = sd(value))

ggplot(transp_var, aes(x= factor(transp), y = value, fill = factor(transp)))+
  geom_boxplot() +
  scale_fill_brewer(palette = "YlGn") + # Change fill colors
  theme_minimal() + # Use a minimal theme
  labs(title = "Transportation Indicators by Neighborhood Profiles", 
       x = "Cluster", 
       y = "Value",
       fill = "cluster") +
  facet_wrap(~indicator, scales = "free", nrow = 1)
```

It looks like bus-related variables are more influential than the BRT ones in the modeling. For example, this look like this in the map:

Suggesting that cluster 1 is corresponding to neighborhoods with more bus routes and density.
