---
title: "Street Profiles for Bogotá Neighborhoods (ZAT)"
subtitle: "Finite Mixture Modeling with `flexmix`"
author: "Heli Xu"
date: 1/31/24
format: 
  html:
    code-fold: true
    toc: true
    toc-depth: 4
    self-contained: true
editor: visual
---

After checking distribution of each variable at ZAT level (with mean and variance, see [shiny dashboard](https://heli-xu.shinyapps.io/zat-indicator/)), we'll now try using finite mixture modeling to fit our data to understand the neighborhood profiles.

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(tibble)
library(flexmix)
library(ggbeeswarm)

```

## Correlation between variables

For count data, Poisson and Negative Binomial distribution are commonly used for modeling. They generally assume variables are independent of each other, we'll first explore the correlation among these variables.

```{r}
zat_std2 <- readRDS("../../../../clean_data/zat_std2.rds")
zat_std3 <- readRDS("../../../../clean_data/zat_std3.rds")

zat_cor <- zat_std2 %>% select(-ZAT) %>% cor()
zat_cor <- zat_std3 %>% select(-ZAT) %>% cor()

zat_cor[lower.tri(zat_cor, diag = FALSE)] <- NA
# >0.6 related variables
zat_cor
```

The variables that show strong correlation are:

```{r}
as.data.frame(zat_cor) %>%
  rownames_to_column("var1") %>% 
  pivot_longer(cols = -var1, names_to = "var2", values_to = "cor") %>% 
  filter(cor >0.6 & cor != 1)
```

Therefore, we'll exclude `st_4ln_length_log`, `INTDENS`, `NUMRBP`, and `NUMRT` in our modeling.

INTDENS, LONGMV, NUMRBP, NUMRT, LONGRBP, LONGRT

```{r, message=FALSE}

street <- zat_std2 %>% 
  select(road_length_log, NUMINT, trlight_per_int, tree_per_km2) %>% 
  colnames()

street <- zat_std3 %>% 
  select(LRDENS, NUMINT, trlight_per_int) %>% 
  colnames()

transportation <- zat_std2 %>% 
  select(BUSTOPDENS, bus_length_log,brt_length_log, BPRDRATE, bridg_per_km2) %>% 
  colnames()

all <- zat_std2 %>% 
  select(-c(ZAT, st_4ln_length_log, INTDENS, NUMRBP, NUMRT)) %>% 
  colnames()

#binary
fmm_binary <- function(domain, k){
  var_to_model <- zat_std2 %>% select(-ZAT, all_of(domain))
  
  mix <- stepFlexmix(as.matrix(var_to_model) ~ 1, data = zat_std2, 
  model = FLXMCmvbinary(), k = k, 
  nrep = 3)
  
  return(mix)
}

street_mix<- fmm_binary(street, 1:7)
transp_mix <- fmm_binary(transportation, 1:7)
all_mix <- fmm_binary(all, 1:7)

elbow_bic <- function(mix, k){
  bic <- BIC(mix)
  result <- data.frame(Clusters = k, BIC = bic)
  
  ggplot(results, aes(x = Clusters, y = BIC)) +
  geom_line(color = "blue", ) +
  geom_point() +
  theme_minimal() +
  labs(title = "Elbow Plot for BIC Values",
    x = "Number of Clusters",
    y = "BIC")
}

elbow_bic(street_mix, 1:7)
elbow_bic(transp_mix, 1:7)

#beeswarm
mix_best <- getModel(street_mix, "BIC")
mix_best2 <- getModel(transp_mix, "BIC")

plot_profile <- function(mix, domain){
  mix_best <- getModel(mix, "BIC")
  data <- zat_std3 %>% select(ZAT, all_of(domain)) %>% 
    mutate(cluster = clusters(mix_best)) %>% 
    pivot_longer(-c(ZAT, cluster), names_to = "indicator")
  
  data %>% 
    ggplot(aes(x = value, y= indicator, color = factor(cluster)))+
  geom_beeswarm()+
  facet_wrap(~indicator, scales = "free", ncol = 1)+
    theme(
      strip.text = element_blank()
    )
}

plot_profile(street_mix, street)
plot_profile(transp_mix, transportation)
plot_profile(all_mix, all)


```

```{r}
#poisson
fmm_poisson <- function(domain, k){
  var_to_model <- zat_std3 %>% select(-ZAT, all_of(domain))
  
  mix <- stepFlexmix(as.matrix(var_to_model) ~ 1, data = zat_std3, 
  model = FLXMCmvpois(), k = k, 
  nrep = 3)
  
  return(mix)
}

street_pmix <- fmm_poisson(street, 1:7)
transp_pmix <- fmm_poisson(transportation, 1:7)
all_pmix <- fmm_poisson(all, 1:7)

mix_best <- getModel(street_pmix, "BIC")

plot_profile(street_pmix, street)
plot_profile(transp_pmix, transportation)
plot_profile(all_pmix, all)
```

```{r}
library(NB.Clust)
```
