---
title: "Street Profiles for Bogotá Neighborhoods (ZAT)"
subtitle: "Finite Mixture Modeling with `flexmix`"
author: "Heli Xu"
date: 1/31/24
format: 
  html:
    code-fold: true
    toc: true
    toc-depth: 4
    self-contained: true
editor: visual
---

After visualizing ZAT-level variable distributions spatially on maps and statistically through histograms, we'll now try using finite mixture modeling to fit our data to understand the neighborhood profiles.

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(tibble)
library(flexmix)
library(ggbeeswarm)

```

## Data preparation

In the previous post, we used the raw data to explore the variable distributions. To get the data ready for modeling, we'll need to clean and standardize the data.

The cleaning step is straightforward, where we leave out the ZAT units with zero road network length (`LRDENS`, in meters) and/or intersections. 4 ZAT units don't have measurable road network length or intersections, and another 9 ZAT units don't have intersections, which reduce the dataset from 919 rows (ZAT units) to 906 rows.

To standardize the data, we're using `areakm2` to normalize the countinuous and count data that have not already been area-adjusted (eg., `NUMRBP`, `NUMTTFLIGH`, `NUMSTTREES`), and taking a natural log of the length measurements in meters (eg., `LRDENS`, `LONGRBP`), with the bike lane rate transformed into log length of bike lanes.

```{r, eval=FALSE}
zat_std <- zat_raw_data %>%
  filter(LRDENS > 0, 
         NUMINT >0) %>% 
  mutate(
    road_length_log = log(LRDENS),
    st_4ln_length_log = log(LONGMV),
    bikelane_per_km2 = BPRDRATE * LRDENS,
    bikelane_m_log = case_when(bikelane_per_km2 > 0 ~ log(bikelane_per_km2),
                              .default = bikelane_per_km2),
    sttree_per_km2 = NUMSTTREES / areakm2,
    bridg_per_km2 = NUMBRIDGES / areakm2,
    trlight_per_km2 = NUMTTFLIGH / areakm2,
    numrbp_per_km2 = NUMRBP/areakm2,
    numrt_per_km2 = NUMRT/areakm2,
    longrbp_per_km2 = LONGRBP / areakm2,
    longrt_per_km2 = LONGRT / areakm2,
    bus_length_log = case_when(longrbp_per_km2 > 0 ~ log(longrbp_per_km2),
                               .default = longrbp_per_km2),
    brt_length_log = case_when(longrt_per_km2 > 0 ~ log(longrt_per_km2),
                               .default = longrt_per_km2)
  ) 
```

The histograms of the distribution of the original variables, standardized variables and their mean/variance can be viewed on this interactive [shiny dashboard](https://heli-xu.shinyapps.io/zat-indicator/).

We'll select the standardized variables for modeling.

```{r, eval=FALSE}
zat_std2 <- zat_std %>% 
  select(ZAT, BUSTOPDENS, road_length_log, st_4ln_length_log, bikelane_m_log, 
         sttree_per_km2, bridg_per_km2, trlight_per_km2, numrbp_per_km2,
         numrt_per_km2, bus_length_log, brt_length_log)
```

## Correlation among variables

first explore the correlation among these variables.

```{r}
zat_std2 <- readRDS("../../../../clean_data/ZAT/zat_std2.rds")

zat_cor <- zat_std2 %>% select(-ZAT) %>% cor()

zat_cor[lower.tri(zat_cor, diag = FALSE)] <- NA
# >0.6 related variables
zat_cor %>% as_tibble()
```

The variables that show strong correlation are:

```{r}
as.data.frame(zat_cor) %>%
  rownames_to_column("var1") %>% 
  pivot_longer(cols = -var1, names_to = "var2", values_to = "cor") %>% 
  filter(cor >0.6 & cor != 1)
```

The length of toad network is shown to be correlated with both the length of streets with 4 or more lanes and the length of bus routes. Although the number of BRT routes are expectedly correlated with the length of the BRT routes, that's not the case for bus routes. Based on this, we'll exclude `road_length_log` and `numrt_per_km2` in our modeling.

```{r, message=FALSE}

zat_fmm <- zat_std2 %>% select(-road_length_log, -numrt_per_km2)

street <- zat_fmm %>% 
  select(st_4ln_length_log, bikelane_m_log, trlight_per_km2, sttree_per_km2, bridg_per_km2) %>% 
  colnames()

transportation <- zat_fmm %>% 
  select(BUSTOPDENS, bus_length_log, brt_length_log, numrbp_per_km2) %>% 
  colnames()

all <- zat_fmm %>% 
  select(-ZAT) %>% 
  colnames()

#normal
fmm_normal <- function(domain, k){
  var_to_model <- zat_fmm %>% select(-ZAT, all_of(domain))
  
  mix <- stepFlexmix(as.matrix(var_to_model) ~ 1, data = zat_fmm, 
  model = FLXMCmvnorm(diagonal = FALSE), k = k, 
  nrep = 3)
  
  return(mix)
}

street_mix<- fmm_normal(street, 1:7)
transp_mix <- fmm_normal(transportation, 1:7)
all_mix <- fmm_normal(all, 1:7)

elbow_bic <- function(mix, k){
  bic <- BIC(mix)
  results <- data.frame(Clusters = k, BIC = bic)
  
  ggplot(results, aes(x = Clusters, y = BIC)) +
  geom_line(color = "blue", ) +
  geom_point() +
  theme_minimal() +
  labs(title = "Elbow Plot for BIC Values",
    x = "Number of Clusters",
    y = "BIC")
}

elbow_bic(street_mix, 1:7)
elbow_bic(transp_mix, 1:7)

#beeswarm
mix_best <- getModel(street_mix, "BIC")
mix_best2 <- getModel(transp_mix, "BIC")

plot_profile <- function(mix, domain){
  mix_best <- getModel(mix, "BIC")
  data <- zat_fmm %>% select(ZAT, all_of(domain)) %>% 
    mutate(cluster = clusters(mix_best)) %>% 
    pivot_longer(-c(ZAT, cluster), names_to = "indicator")
  
  data %>% 
    ggplot(aes(x = value, y= indicator, color = factor(cluster)))+
  geom_beeswarm()+
  facet_wrap(~indicator, scales = "free", ncol = 1)+
    theme(
      strip.text = element_blank()
    )
}

plot_profile(street_mix, street)
plot_profile(transp_mix, transportation)
plot_profile(all_mix, all)


```

```{r}
#poisson
fmm_poisson <- function(domain, k){
  var_to_model <- zat_std3 %>% select(-ZAT, all_of(domain))
  
  mix <- stepFlexmix(as.matrix(var_to_model) ~ 1, data = zat_std3, 
  model = FLXMCmvpois(), k = k, 
  nrep = 3)
  
  return(mix)
}

street_pmix <- fmm_poisson(street, 1:7)
transp_pmix <- fmm_poisson(transportation, 1:7)
all_pmix <- fmm_poisson(all, 1:7)

mix_best <- getModel(street_mix, "BIC")

plot_profile(street_pmix, street)
plot_profile(transp_pmix, transportation)
plot_profile(all_pmix, all)
```

```{r}
post_p <- posterior(mix_best) %>% as.data.frame() %>% 
  rownames_to_column(var = "ZAT") %>% 
  mutate(order = V1-V2) %>% 
  pivot_longer(c(V1, V2), names_to = "cluster", values_to = "prob")

ggplot(post_p, aes(x = reorder(ZAT, order), y = prob, fill = cluster)) +
  geom_bar(stat = "identity", position = "fill", width = 1) +  # Adjust the width to fill the space
  scale_fill_brewer(palette = "YlGnBu") +
  coord_flip() +  # Rotate the plot to make it look like a heatmap
  scale_x_discrete(expand = c(0, 0)) +  # Remove space between groups
  labs(x = "ZAT Units", y = "Posterior Probabilities", title = "Posterior Probabilities of Cluster Assignment of Each ZAT Unit") +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        title = element_text(face = "bold", size = 13),
        axis.title.x = element_text(size = 13))
```
